---
title: "Colchester Policing and Climate Analysis 2023"
author: "Rakshitha Kalyan Rao"
date: "`r Sys.Date()`"
output:
  html_document:
    toc: true
    toc_float: true
---

```{r setup, include=FALSE}
knitr::opts_chunk$set(echo = TRUE, warning = FALSE, message = FALSE)
```

## 1. Introduction

This report presents an analysis of street-level crime data and daily climate data for Colchester in 2023. The objectives include data visualization, trend analysis, and insights generation.

## 2. Data Loading and Cleaning

The necessary datasets, temp2023.csv for temperature data and crime23.csv for crime occurrences, are imported first in the study. Context and location_subtype, two columns with a high percentage of missing data, were removed from the crime dataset. To make temporal analysis easier, the date column was transformed into a standard Date format. The Date column was reformatted for uniformity, and superfluous columns such as PreselevHp and SnowDepcm were eliminated for the temperature dataset. The datasets are guaranteed to be clean, organised, and prepared for analysis using these preparation procedures. By reducing noise, this seminal study enhances the precision of later visualisations and interpretations.

### Load Libraries

```{r}
#loading required libraries
library(tidyverse)
library(lubridate)
library(leaflet)
library(ggplot2)
library(plotly)
library(GGally)
```

### Loading of Datasets

```{r}
# Load crime data
crime_data <- read.csv("/Users/rakshithakalyanrao/Downloads/crime23[58].csv")

# Load temperature data
temp_data <- read.csv("/Users/rakshithakalyanrao/Downloads/temp2023.csv")
```

### Explore and Clean Data
The datasets is been analysed to determine their quality, substance, and structure. To find data that is inconsistent or missing, summary statistics were produced.Due to the large percentage of missing values in the crime dataset, superfluous columns like context and location_subtype were eliminated. To make time-series analysis easier, the date column was also converted into a standard Date type.Non-essential variables such as PreselevHp and SnowDepcm were removed from the temperature dataset. To maintain uniformity between analyses, the Date column was also changed to a Date object.By streamlining the data, these preparation procedures made sure it was clear and prepared for insightful analysis and visualisation.


#### Crime Data

```{r}
# View structure and summary
str(crime_data)
summary(crime_data)

# Drop columns with excessive missing data
crime_data <- crime_data %>% select(-context, -location_subtype)

# Convert date column to Date format
crime_data$date <- as.Date(paste0(crime_data$date, "-01"), format = "%Y-%m-%d")
```

#### Temperature Data

```{r}
# View structure and summary
str(temp_data)
summary(temp_data)

# Drop columns with excessive missing data
temp_data <- temp_data %>% select(-PreselevHp, -SnowDepcm)

# Convert Date column to Date format
temp_data$Date <- as.Date(temp_data$Date, format = "%Y-%m-%d")
```

## 3. Data Visualizations

### 3.1 Tables

#### Frequency Table of Crime Categories
To compile the prevalence of different crime types in the Colchester dataset, a frequency table was made. The distribution of crimes is succinctly summarised in this table, which aids in determining the most and least prevalent categories.

keypoints:
The table shows which crime categories account for the largest portion of the total count.
Prioritising resources for law enforcement or community safety programs might be made easier with an understanding of these distributions.By emphasising the main crime categories and their frequency in the dataset, this preliminary investigation establishes the groundwork for additional investigation.

```{r}
#Frequency Table
crime_table <- table(crime_data$category)
knitr::kable(as.data.frame(crime_table), col.names = c("Category", "Count"))
```

### 3.2 Bar Plot

```{r}
#Bar plot of crime data
ggplot(crime_data, aes(x = category, fill = category)) +
  geom_bar() +
  theme(axis.text.x = element_text(angle = 45, hjust = 1)) +
  labs(title = "Crime Categories in Colchester", x = "Category", y = "Count", fill = "Crime Category")
```

### 3.3 Histogram of Temperatures

```{r}
#Histogram of temp data
ggplot(temp_data, aes(x = TemperatureCAvg, fill = ..count..)) +
  geom_histogram(binwidth = 1, color = "white") +
  scale_fill_gradient("Count", low = "blue", high = "red") +
  labs(title = "Histogram of Average Temperatures", x = "Temperature (°C)", y = "Frequency")
```

### 3.4 Time Series Plot

```{r}
#Time Series plot for temp data
ggplot(temp_data, aes(x = Date, y = TemperatureCAvg)) +
  geom_line(color = "red") +
  geom_smooth(method = "loess") +
  labs(title = "Average Temperature Over Time", x = "Date", y = "Temperature (°C)")
```

### 3.5 Interactive Map of Crime Locations

```{r}
leaflet(data = crime_data) %>%
  addTiles() %>%
  addCircleMarkers(~long, ~lat, radius = 3, color = "blue", popup = ~paste("Category:", category)) %>%
  setView(lng = mean(crime_data$long), lat = mean(crime_data$lat), zoom = 12) %>%
  addLegend(position = "bottomright", colors = "blue", labels = "Crime Locations", title = "Legend")
```

### 3.6 Correlation Analysis

#### Correlation Between Temperature Variables

```{r}
# Calculate correlation matrix
temp_corr <- temp_data %>% select(TemperatureCAvg, TemperatureCMax, TemperatureCMin) %>% cor()

# Display correlation matrix
temp_corr

# Plot correlation matrix
ggplotly(ggcorr(temp_data %>% select(TemperatureCAvg, TemperatureCMax, TemperatureCMin), label = TRUE))
```
